---
title: '\small Introduction to "R"\newline\normalsize Data wrangling'
author: "Arndt Leininger\\newline\\newline \\faTwitter \\href{https://twitter.com/a_leininger}{@a\\_leininger} \\newline \\Email \\href{mailto:arndt.leininger@fu-berlin.de}{arndt.leininger@fu-berlin.de}"
date: "28 September 2019"
output:
  beamer_presentation:
    toc: false
    colortheme: "seahorse"
    fonttheme: "structurebold"
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
options(width=55)
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, cache = TRUE,
                      tidy=TRUE, tidy.opts=list(width.cutoff=55))
setwd('~/Git/r_workshop/slides')
```

# Installing packages

## Installing packages

- In the following section we will use the packages `dplyr` and `tidyr`
- These packages were created by Hadley Wickham and are available on CRAN
- They are not shipped with R so you need to install them before loading and using them
- `install.packages()` takes a package name as input and installs the requisite package

```{r}
install.packages('dplyr')
install.packages('tidyr')
```

or

```{r}
install.packages(c('dplyr', 'tidyr'))
```

- Package names must be passed on as a character vector

## Loading packages

- Before you can use the functions provided by the new packages you need to load them

```{r library3, eval=TRUE}
library('dplyr')
library(tidyr)
```

## Duplicate function names

Sometimes two different functions have the same name

```{r lag_example, eval=TRUE}
lag(c(1,2,3))

stats::lag(c(1,2,3))
```

## ::

Using `[package]::[function]` you can use a function without first loading the package with `library`

```{r, eval=TRUE}
stringr::str_replace(c('Hello', 'World', '!'), '!', '.')
```

## Quick hands-on

1. Install the packages `dplyr` and `tidyr`.
2. Load `dplyr` and `tidyr`.
3. Create a vector containing the numbers `0, 1, 1, 2, 3, 1, 0, 0`
4. Use the function `lag()` on that vector.
5. use `dplyr::lag()` and `stats:lag()` on the vector. Which one is used if you use `lag()`?

## *Beware of `require()`*

- Don't use `require()` which does not provide an error message if a package could not be loaded

```{r require, eval=TRUE, error=TRUE}
require('thispackagedoesnotexit')
library('thispackagedoesnotexit')
```

https://yihui.name/en/2014/07/library-vs-require/

## *Installing development versions*

```{r devtools}
install.packages('devtools')

devtools::install_github('bookdown')
```

# Data wrangling

## Data wrangling

- Remember that some say that quantitative social science research is 80% data collection, cleaning and wrangling and 20% actual analysis.
- Data don't always come in nicely formatted files (e.g. .dta). Sometimes you have a data source or sources in files that aren’t set up for R to read and start analyzing right away.
- You often have to do a bit of work to get things into the format you want: the more of that work is recorded the better.
- `R` and the packages `dpylr` and `tidyr` amongst others help you reduce the time you spend wrangling and makes that time more fun

## dplyr

- `dplyr` is a package which simplifies data wrangling a lot
- It provides a number of functions which together allow a wide range of data manipulation
    - `arrange()`: sorting data
    - `filter()`: subsetting data 
    - `select()` (and `rename()`): subsetting data (by column)
    - `mutate()`: add or replace variables
    - `group_by()`: do a `mutate()` or `summarize()` group-wise
    - `summarize()`: aggregate data
- It provides 'piping' functionality


## Wrangling

Sort domestic cars by price

```{r readdta_hidden, eval=TRUE, echo=FALSE}
library(foreign)
df <- read.dta('../data/auto.dta')
```

```{r, eval=TRUE}
df %>% filter(foreign == 'Domestic') %>% arrange(price) %>% 
    select(make, foreign, price)
```

## Piping

- `%>%` is a 'piping' operator, it passes on output from a function to another, allowing you to write code in the order of execution
- The package `dpylr` provides piping functionality for R through the package `magrittr`

```{r pipe_example, eval=TRUE}
df %>% filter(price == max(price)) %>% select(make, price)
# is easier to read than
select(filter(df, price == max(price)), make, price)
```

Keyboard shortcut \keys{\ctrl + \shift + M} produces the piping symbol

## Wrangling
The average price of domestic cars is also simple to obtain

```{r subsetting2, eval=TRUE, tidy=FALSE}
df %>% filter(foreign == 'Domestic') %>% 
  summarise(average = mean(price))
```

Read more about it at http://blog.revolutionanalytics.com/2014/07/magrittr-simplifying-r-code-with-pipes.html

## Creating a new variable

Create a variable which indicates whether a car's price is above average, distinguishing between domestic and foreign cars.

```{r wrangling3, eval=TRUE, tidy=FALSE}
tmp <- df %>% group_by(foreign) %>% 
  mutate(relativeprice = price - mean(price),
         expensive = relativeprice > 0)
tmp %>% select(price, relativeprice, expensive)
```

## Aggregating data

- `summarize()` in conjunction with `group_by()` can be used to aggregate data
- function used within `summarize()` must return a single value

```{r summarize, eval=TRUE}
df %>% group_by(foreign) %>% summarize(price = mean(price))
```


## A tibble

- `as_tibble()` turns a data.frame into a slightly modified data.frame (a 'tibble') with improved printing capabilities

```{r as_tibble, eval=TRUE}
df <- as_tibble(df)
df
```

## A tibble

- `as_tibble()` turns a data.frame into a slightly modified data.frame with improved printing capabilities

```{r as_tibble2, eval=TRUE}
df <- as_tibble(df)
print(df, n = 5)
```

## `glimpse()`

```{r glimpse, eval=TRUE}
glimpse(df)
```


## Tidyverse

```{r tidyverse}
install.packages('tidyverse')
library(tidyverse)
```

- collection of useful R [packages](https://www.tidyverse.org/packages/) by [Hadley Wickham](http://hadley.nz/) (e.g. dplyr, tidyr, ggplot2, readr, stringr)
- https://www.tidyverse.org/


# Hands-on I

## Hands-on I

`hands-on/03_datawrangling/hands-on1.R`

# Merging and reshaping data

## Merging data

- `dplyr` provides great functions for merging data
- `left_join` keeps all observations in `d1` and adds those in `d2` that can be matched
- Takes two data.frames as input
- Identifier variable names can differ

```{r merge}
d1 <- read.csv('data/d1.csv')
d2 <- read.csv('data/d2.csv')
m <- left_join(d1, d2, by = c('nation' = 'country', "year"))
```

- there are also: `inner_join`, `right_join`, `semi_join`,
  `anti_join`, `full_join`

## Tidying data

- Tidy data refers to data in which each row is an observation and each column a variable
- This is also known as 'long' format as opposed to 'wide' format
- Real-world data do not always come in this form
- The package `tidyr` provides functions to 'clean up' data

```{r, echo=FALSE, eval=TRUE}
library(tidyr)
```

```{r}
install.packages('tidyr')
library(tidyr)
```

## Reshaping data

- Let's use a very simple made-up `data.frame` 

```{r messy, eval=TRUE}
messy <- data.frame(
  name = c("Wilbur", "Petunia", "Gregory"),
  a = c(67, 80, 64),
  b = c(56, 90, 50)
)

messy
```

- Imagine this is data from a clinical trial. We have three variables, patient name, drug and heartrate, but only one is a variable yet
- Based on https://blog.rstudio.org/2014/07/22/introducing-tidyr/

## Reshaping data

- We will use `tidyr`'s `gather()` to gather the a and b columns into key-value pairs of `drug` and `heartrate`

```{r gather, eval=TRUE}
tidy <- gather(messy, key = drug, value = heartrate, a:b)

tidy
```

## And back to wide format

```{r spread, eval=TRUE}
spread(tidy, drug, heartrate)
```

# Hands-on II

## Hands-on II

`hands-on/03_datawrangling/hands-on2.R`