---
title: '\small \texttt{R} Statistical Software in Action for Newcomers\newline\normalsize Data Analysis'
author: "Arndt Leininger\\newline\\newline \\faTwitter \\href{https://twitter.com/a_leininger}{@a\\_leininger} \\newline \\Email \\href{mailto:arndt.leininger@fu-berlin.de}{arndt.leininger@fu-berlin.de}"
date: "24 April 2018"
output:
  beamer_presentation:
    toc: false
    colortheme: "seahorse"
    fonttheme: "structurebold"
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
options(width=55)
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, 
                      tidy=TRUE, tidy.opts=list(width.cutoff=55), cache = TRUE)
# setwd('~/Git/r_workshop/slides')
library(magrittr)
```

## Program

**Tuesday, 24 April 2018**

09:00h - 10:30h Data Analysis

*10:30h - 10:45h break*

10:45h - 12:15h Data visualization

12:15h - 12:30h Concluding remarks

# Correlation and difference-in-means

## Correlation

```{r cor, eval=TRUE}
library(foreign)
d <- read.dta('../data/EUsuppDK.dta')

cor(d$age, d$left_right)

cor(d$age, d$left_right, use = 'complete.obs')
```


## T-Test

```{r ttest, eval=TRUE}
t.test(left_right ~ sex, d)
```

- Missings (`NA`) are dropped automatically

## T-Test

```{r ttest2, eval=TRUE}
t.test(left_right ~ sex, d, var.equal = T)
```

# Simple regression models

## Regression

- Regression models are functions that are fed an equation and data
- Further options are possible but optional
- The dependent variable is seperated by a tilde from the independent variables
- Equation: `dv ~ iv`
- No need for `$` operator in the equation

```{r regression, eval=TRUE}
lm(left_right ~ age, d)
```

## Formulae

**A complete overview of formulae in R**

https://ww2.coastal.edu/kingw/statistics/R-tutorials/formulae.html

## Model objects

- You can save the output of the `lm()` function just like with any other function.

```{r m_summary, eval=TRUE}
m1 <- lm(left_right ~ age, d)
```

- `lm()` is for linear models, i.e. OLS

## Model objects

- Once you save an estimated model as object you can always access it to obtain model statistics.

```{r functions}
summary(m1)  # estimation results

coef(m1)  # coefficients

vcov(m1)  # Variance-Covariance Matrix

predict(m1)  # Predicted values

resid(m1)  # Residuals
```

## Model objects

```{r summary2, eval=TRUE}
summary(m1)  # estimation results
```

## A little trick

```{r summary_m, eval=TRUE}
summary(m <- lm(left_right ~ age, d))
```

## A little trick

```{r summary_m2, eval=TRUE, error=TRUE}
# but
summary(m = lm(left_right ~ age, d))
```

## Model objects

```{r coef, eval=TRUE}
coef(m1)  # coefficients
```

## Model objects

```{r vcov, eval=TRUE}
vcov(m1)  # Variance-Covariance Matrix
```

## Model objects

```{r predict, eval=TRUE}
predict(m1)  # Predicted values
```

## Model objects

```{r residuals, eval=TRUE}
resid(m1)  # Residuals
```


## Model objects

```{r objects, eval=TRUE}
class(m1)
objects(m1)
```

## Interactions

- Interactions can be specified as follows
    - `var1*var2` = `var1 + var2 + var1:var2 `
    - `var1:var2` is simply the interaction term
    
    
```{r, eval=TRUE}
m_i <- lm(left_right ~ sex*age, d)
# or
m_i <- lm(left_right ~ sex + age + sex:age, d)
```

## Polynomials

```{r}
lm(left_right ~ age + I(age^2), d) # second-order polynomial

lm(left_right ~ age + age^2, d) # does not work
```

- `?I()`: "Change the class of an object to indicate that it should be treated 'as is'."

## A little trick

- Put the assignment of a model to an object in a `summary()` call to assign and view results at the same time

```{r, eval=TRUE}
summary(m2 <- lm(left_right ~ age + I(age^2), d))
```

## Time-Series analysis

- Lags and leads can be easily included in time-series data
- For this make sure that the data are sorted on the time variable
    - Remember to use `order()` not `sort()``(which is for vectors) to sort a data.frame
    - Or use `dplyr`'s `arrange()`
- `lag()` and `lead()`
- For second lag: `lag(var, 2)`

```{r}
summary(m_l <- lm(y ~ x + lag(x) + lag(x, 2), d))
```

## Logistic regression

- Generalized linear models, such as probit or logistic regression, are provided through the `glm()` function

```{r}
d$wealthy <- ifelse(d$wealth == '++' | d$wealth == '+', T, F)

# Probit
m_g <- glm(wealthy ~ age + sex, data = d)

# Logistic
m_g <- glm(wealthy ~ age + sex, family = binomial(), data = d)
```

## Predicted values

- `predict()` is a generic function to create predictions from various models

```{r yhat, eval=TRUE, error=TRUE}
predicted_values <- predict(m1)
# but...
d$yhat <- predict(m1)
# Error in `$<-.data.frame`(`*tmp*`, "yhat", 
# value = c(5.41494877919278,  : 
# replacement has 968 rows, data has 1001
```

## Predicted values

```{r yhat2, eval=TRUE, error=TRUE}
d$yhat <- predict(m1, newdata = d)
# no error message because now predictions 
# are also made for deleted observations;
# these predictions are obviously NA
```

## Predicted values

Or...

```{r yhat3, error=TRUE}
install.packages('broom')
```

```{r yhat4, eval=TRUE, error=TRUE}
library(broom)
augment(m1) %>% head
```


## Predicted values

```{r yhat5, eval=TRUE, error=TRUE}

df <- data.frame(age = min(d$age):max(d$age))

predict(m1, newdata = df)

```


# Hands-on I

## Hands-on I

https://gitlab.com/arndtl/r_workshop

# Presenting Results with R

## Lists

- Lists can contain any kind of objects of any type.
- Note: data.frames can also contain vectors of any of the three types but the vectors are forced to be of the same length.
- Example: One could have a list of differently sized vectors

```{r lists, eval=TRUE}
v1 <- c(1, 2, 3)
v2 <- c('a', 'b', 'c', 'd')
alist <- list(v1, v2)
alist
```

## Tables

Using packages such as `stargazer` or `texreg` we can create nice regression tables.

```{r, eval=TRUE, message=TRUE, echo=TRUE, warning=TRUE}
library(stargazer)
library(texreg)
```

## Tables

Here's an example in `tex` using `stargazer`.

```{r lm, results='asis', eval=TRUE, tidy=FALSE}
m1 <- lm(mpg ~ cyl, mtcars)
m2 <- lm(mpg ~ cyl + gear, mtcars)

stargazer(list(m1, m2), header = F, float = F, 
          font.size = 'tiny', single.row = T)
```

## Tables

`stargazer` provides tables in `text`, `html` and `tex.` 

```{r lm2}
# output as text file
stargazer(m1,  type = 'text', out = 'tables/m1.txt')
# output as html file which Word can read
stargazer(m1, type = 'html', out = 'tables/m1.html')
# output as tex, the default
stargazer(m1, out = 'tables/m1.tex')
```

## Screenreg

- `texreg`'s `screenreg()` function is very useful to quickly view some models.

```{r screenreg, eval=TRUE}
screenreg(list(m1, m2))
```


## Tables

`texreg` provides tables `html`, `tex` and to screen. 

```{r lm3}
# output as tex file
texreg(m1,  file = 'tables/m1.tex')
# output as html file which Word can read
htmlreg(m1, file = 'tables/m1.html')
# output as tex, the default
screenreg(m1)
```


## Regression tables for Word

**Via HTML**

- Export to HTML using `htmlreg()` (package `texreg`) or `stargazer(...,  type = 'html')`
- Then copy and paste to Word
- Or, better, link to the html file from within the Word Document
    - Word: Insert -> Object (dropwdown) -> Text from File -> Insert (dropdown) -> Insert as link; hit F9 to refresh
    - LibreOffice Write: Insert -> Section -> Check option "Link" and choose document; to refresh: Edit -> Links -> click "Update"     
    - http://www.techrepublic.com/article/link-to-another-file-in-your-word-document/

## Important packages for regression analysis

- `lmtest` for F-test and other tests
- `tseries` for time-series analysis
- `plm` for panel data analysis
- `lme4` for multilevel models
- `MatchIt` and `Matching` for matching

# Hands-on II

## Hands-on II

https://gitlab.com/arndtl/r_workshop

# Further packages

## ReporterRs

- `ReporterRs` is a package allows creation of entire(!) Word and Power Point documents
  - this includes (regression) tables

```{r install_reporteRs}
install.packages('ReporteRs')
```

http://davidgohel.github.io/ReporteRs/index.html

## Regression tables with ReporterRs

```{r reporters, message=FALSE, tidy=FALSE}
library(ReporteRs)

# save the model in a data.frame
mdata <- as.data.frame(summary(m1)$coefficients)

# Define significance cutoffs
signif.codes <- cut(mdata[,4], 
                    breaks = c( -Inf, 0.001, 0.01, 0.05, Inf), 
                    labels= c("***", "**", "*", "" ) )

#format the values of coefficients, etc.
mdata[, 1:3] <- apply(mdata[, 1:3], 2, round, 2)
mdata[, 4] <- ifelse(mdata[, 4] < .001, "< 0.001", 
                     round(mdata[, 4], 3))
# add signif codes to data
mdata$Signif = signif.codes
```

## Regression tables with ReporterRs

```{r reporters2, tidy=FALSE}
mdata

# create an empty FlexTable
coef_ft = FlexTable(data = mdata, add.rownames=TRUE, 
                    body.par.props = parRight(), 
                    header.text.props = textBold(),
                    header.columns = T)
```

## Regression tables with ReporterRs

```{r reporters3, tidy=FALSE}
# format the table a bit
coef_ft = setFlexTableBorders(coef_ft, 
                              inner.vertical = borderNone(), 
                              inner.horizontal = borderNone(), 
                              outer.vertical = borderNone(), 
                              outer.horizontal = borderSolid())
```

## Regression tables with ReporteRs

Now we'll save the table in a Word document.

```{r reporters4}
# Create an empty document
doc <- docx()

# Add the regression table to the document
doc <- addFlexTable(doc, coef_ft)

# Save the document to the hard drive
writeDoc(doc, file = '../tables/regtable.docx')
```