---
title: '\small Introduction to "R"\newline\normalsize Recap and model estimation'
author: "Arndt Leininger\\newline\\newline \\faTwitter \\href{https://twitter.com/a_leininger}{@a\\_leininger} \\newline \\Email \\href{mailto:arndt.leininger@fu-berlin.de}{arndt.leininger@fu-berlin.de}"
date: "29 September 2019"
output:
  beamer_presentation:
    toc: false
    colortheme: "seahorse"
    fonttheme: "structurebold"
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
# setwd('~/Git/r_workshop/slides')
options(width=55)
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, cache = FALSE,
                      warning = FALSE, message = FALSE,
                      tidy=TRUE, tidy.opts=list(width.cutoff=55),
                      dev = 'png',
                      fig.width=12, fig.height=7)
library(magrittr)
library(ggplot2)
library(ggthemes)
library(dplyr)

theme_set(theme_gray(base_size = 20))
```

# Exercise and recap

## Exercise and recap

**Good morning!**

Continue your work on `hands-on/03_datawrangling/hands-on1.R`.

If you're done, you can begin solving the take-home exercise:
`exercise/R_exercise_1_questions.pdf`. You can write your solutions into
`exercise/R_exercise_1.R`.

\includegraphics[width=.3\textwidth]{img/coffee.png}

## Program

**Sunday, 29 September 2019**

10:00h - 11:15h Recap and Model Estimation

*11:15h - 11:30h break*

11:30h - 13:00h Data Visualization

*13:00h - 14:00h Lunch break*

14:00h - 16:30h Data and Model Visualization

## OLS Regression

- Regression models are functions that are fed an equation and data
- Further options are possible but optional
- The dependent variable is seperated by a tilde from the independent variables
- Equation: `dv ~ iv`
- No need for `$` operator in the equation

```{r regression, eval=FALSE}
library(foreign)
d <- read.dta('../data/EUsuppDK.dta')

lm(left_right ~ age, d)
```

## OLS Regression

```{r regression2, eval=TRUE}
library(foreign)
d <- read.dta('../data/EUsuppDK.dta')

lm(left_right ~ age, d)
```

## Formulae

**A complete overview of formulae in R**

https://ww2.coastal.edu/kingw/statistics/R-tutorials/formulae.html

## Model objects

- You can save the output of the `lm()` function just like with any other function.

```{r m_summary, eval=TRUE}
m1 <- lm(left_right ~ age, d)
```

- `lm()` is for linear models, i.e. OLS

## Model objects

- Once you save an estimated model as object you can always access it to obtain model statistics.

```{r functions}
summary(m1)  # estimation results

coef(m1)  # coefficients

vcov(m1)  # Variance-Covariance Matrix

predict(m1)  # Predicted values

resid(m1)  # Residuals
```

## Model objects

```{r summary2, eval=TRUE}
summary(m1)  # estimation results
```

## Model objects

```{r coef, eval=TRUE}
coef(m1)  # coefficients
```

## Model objects

```{r predict, eval=TRUE}
predict(m1)  # Predicted values
```

## Model objects

```{r residuals, eval=TRUE}
resid(m1)  # Residuals
```

## Model objects

```{r objects, eval=TRUE}
class(m1)
objects(m1)
```

## Interactions

- Interactions can be specified as follows
    - `var1*var2` = `var1 + var2 + var1:var2 `
    - `var1:var2` is simply the interaction term
    
```{r, eval=TRUE}
m_i <- lm(left_right ~ sex*age, d)
# or
m_i <- lm(left_right ~ sex + age + sex:age, d)
```

## Polynomials

```{r}
lm(left_right ~ age + I(age^2), d) # second-order polynomial

lm(left_right ~ age + age^2, d) # does not work
```

- `?I()`: "Change the class of an object to indicate that it should be treated 'as is'."

## A little trick

- Put the assignment of a model to an object in a `summary()` call to assign and view results at the same time

```{r, eval=TRUE}
summary(m2 <- lm(left_right ~ age + I(age^2), d))
```

## Predicted values

- `predict()` is a generic function to create predictions from various models

```{r yhat, eval=TRUE, error=TRUE}
predicted_values <- predict(m1)
# but...
d$yhat <- predict(m1)
# Error in `$<-.data.frame`(`*tmp*`, "yhat", 
# value = c(5.41494877919278,  : 
# replacement has 968 rows, data has 1001
```

## Predicted values

```{r yhat2, eval=TRUE, error=TRUE}
d$yhat <- predict(m1, newdata = d)
# no error message because now predictions 
# are also made for deleted observations;
# these predictions are obviously NA
```

# Hands-on I

## Hands-on I

`hands-on/04_modelestimation/hands-on1.R`

# Regression tables

## Lists

- Lists can contain any kind of objects of any type.
- Note: data.frames can also contain vectors of any of the three types but the vectors are forced to be of the same length.
- Example: One could have a list of differently sized vectors

```{r lists, eval=TRUE}
v1 <- c(1, 2, 3)
v2 <- c('a', 'b', 'c', 'd')
alist <- list(v1, v2)
alist
```

## Tables

Using packages such as `stargazer` or `texreg` we can create nice regression tables.

```{r, eval=TRUE, message=TRUE, echo=TRUE, warning=TRUE}
library(stargazer)
library(texreg)
```

## Tables

Here's an example in `tex` using `stargazer`.

```{r lm, results='asis', eval=TRUE, tidy=FALSE}
m1 <- lm(mpg ~ cyl, mtcars)
m2 <- lm(mpg ~ cyl + gear, mtcars)

stargazer(list(m1, m2), header = F, float = F, 
          font.size = 'tiny', single.row = T)
```

## Tables

`stargazer` provides tables in `text`, `html` and `tex.` 

```{r lm2}
# output as text file
stargazer(m1,  type = 'text', out = 'tables/m1.txt')
# output as html file which Word can read
stargazer(m1, type = 'html', out = 'tables/m1.html')
# output as tex, the default
stargazer(m1, out = 'tables/m1.tex')
```

## Screenreg

- `texreg`'s `screenreg()` function is very useful to quickly view some models.

```{r screenreg, eval=TRUE}
screenreg(list(m1, m2))
```


## Tables

`texreg` provides tables `html`, `tex` and to screen. 

```{r lm3}
# output as tex file
texreg(m1,  file = 'tables/m1.tex')
# output as html file which Word can read
htmlreg(m1, file = 'tables/m1.html')
# output as tex, the default
screenreg(m1)
```


## Regression tables for Word

**Via HTML**

- Export to HTML using `htmlreg()` (package `texreg`) or `stargazer(...,  type = 'html')`
- Then copy and paste to Word
- Or, better, link to the html file from within the Word Document
    - Word: Insert -> Object (dropwdown) -> Text from File -> Insert (dropdown) -> Insert as link; hit F9 to refresh
    - LibreOffice Write: Insert -> Section -> Check option "Link" and choose document; to refresh: Edit -> Links -> click "Update"     
    - http://www.techrepublic.com/article/link-to-another-file-in-your-word-document/

# Hands-on II

## Hands-on II

`hands-on/04_modelestimation/hands-on2.R`