---
title: '\small Introduction to "R"\newline\normalsize Data manipulation'
author: "Arndt Leininger\\newline\\newline \\faTwitter \\href{https://twitter.com/a_leininger}{@a\\_leininger} \\newline \\Email \\href{mailto:arndt.leininger@fu-berlin.de}{arndt.leininger@fu-berlin.de}"
date: "28 September 2019"
output:
  beamer_presentation:
    toc: false
    colortheme: "seahorse"
    fonttheme: "structurebold"
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
options(width=55)
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, cache = TRUE,
                      tidy=TRUE, tidy.opts=list(width.cutoff=55))
setwd('~/Git/r_workshop/slides')
```

# Reading files and loading packages

## *Reading Data*

Some things to pay attention to when reading data

- Keep an original copy of the data exactly as you found it, if you make changes save to a new name
- Try and make changes with R in a script. If you have to switch to Excel at least write down what you did. ("Friends don't let friends use Excel for data analysis")
- Check the data has been imported properly before you use it. 
    - You may need to specify what character signifies missing values in your data
    - You might need to specify the delimiter in a `.csv` file

## Reading a file from another statistical package

- The package `foreign` provides functions to read datasets saved in other statistical software
- The function we'll use is `read.dta()`
- The package `foreign` is included in every installation of R but is not loaded when starting R
- First we have to load the foreign package

## Loading packages

- R comes with a number of build-in packages and thousands of community contributed packages which extend its functionality
- You can load installed libraries with the `library()` function

```{r library1}
library("foreign")
```

```{r library2, eval=TRUE}
library(foreign)
```

- With the library function there's no need to put the package name in quotation marks

## Reading a (Stata) file

```{r readdta1, eval=TRUE, echo=FALSE}
df <- read.dta('../data/auto.dta')
```


```{r readdta2}
df <- read.dta('data/auto.dta')

```

```{r readdta3, eval=TRUE}
dim(df)
names(df)
```


# Data manipulation

## `data.frame`s

- A `data.frame` is a matrix which allows (column) vectors to be of different types
- A `matrix` can be thought of as a collection of column or row vectors

## Vectors

- Any variable is a vector
- A `data.frame` is simply a collection of vectors of the same length (bot not necessarily type)

```{r vector1, eval=TRUE}
make <- df$make

make
```

## Acccessing elements of a vector

```{r vector2, eval=TRUE}
# First element of a vector
make[1]

# Second element of a vector
df$make[2]

# Elements three to five
make[3:5]

# Elements three and five
df$make[c(1, 3)]
```

## Acccessing elements of a vector

```{r vector 3, eval=TRUE}
price <- as.integer(c(222000000, 120000000, 105000000))

names(price) <- c('Neymar', 'Coutinho', 'Dembélé')

price['Neymar']

price

```

## Enter the matrix

- A `data.frame` is a matrix allowing (column) vectors to be of different types
- A `matrix` can be thought of as a collection of column or row vectors

```{r matrix, eval=TRUE}
# A matrix

m <- matrix(data = c('1,1', '1,2', '1, ...', '1, k',
                '2,1', '2,2', '2, ...', '2, k',
                '...,1', '..., 2', '...', '..., k',
                'n, 1', 'n, 2', 'n, ...', 'n,k'),
       byrow = T, ncol = 4)
m
```

## Enter the matrix

- These can be accessed by stating the row or column index

```{r matrix3, eval=TRUE}
#e.g.
m[1,]  # the first row
m[, 2] # the second column
```

## Matrices

- A rectangle of dimension n (number of rows) and k (number of columns)
- Any element of a matrix can be accessed by reference to its row and column position
- `matrixname[row number(s), column name(s) or number(s)]`

```{r matrix2, eval=TRUE, error=TRUE}
# e.g.
m[1, 2]
m[2, -3]
m[5, 2]
```

## Subsetting

- `data.frame`s can be subset column-wise based on column number(s) or name(s)

```{r col, eval=TRUE}
df[, 1]
```

## Subsetting

- `data.frame`s can be subset column-wise based on column number(s) or name(s)

```{r var, eval=TRUE}
df[, 'make']
```

## Subsetting

- `data.frame`s can be subset column-wise based on column number(s) or name(s)

```{r, eval=TRUE}
df[, c(1, 3)]
```

## Subsetting

- `data.frame`s can be subset column-wise based on column number(s) or name(s)

```{r vars, eval=TRUE}
df[, c('make', 'mpg')]
```

## Subsetting

- `data.frame`s can be subset row-wise based on row number(s) or boolean statements
- Suppose you want to know what the most expensive car is

```{r subset, eval=TRUE}
df[which(df$price == max(df$price)), 'make']
```

## What is `which()`?

- `which()` takes a logical vector as input and returns the index values for `TRUE`

```{r which, eval=TRUE}

d <- data.frame(x = c(1, 2, 3), y = c(21, 18, NA))

d[d$y == 18, ]

d[which(d$y == 18), ]

```

- ignoring `NA`s

## Subsetting

Say you want to subset the data to domestic cars

```{r subset2, eval=TRUE}
df2 <- df[which(df$foreign == 'Domestic'), ]
df2
```

- The comma within the `[]` is mandatory even if you do not subset by column
- `which()` turns the boolean vector into a numeric vector of the indices of `TRUE` values

## Subsetting

The average price of domestic cars is also simple to obtain

```{r subset3, eval=TRUE}
mean(df[which(df$foreign == 'Domestic'), 'price'])
```

## Subsetting

Subset to expensive domestic cars, i.e. cars that are "Domestic" and cost more than 10000

```{r subset5, eval=TRUE}
df2 <- df[which(df$foreign == 'Domestic' & df$price > 10000),]
```

- `&` means AND
- `|` means OR

## Digression: `&` and `|`

What are the results of the following?

1. T & T =
2. T & F =
3. T | T =
4. T | F =

## Digression: `&` and `|`

What are the results of the following?

1. T & T = T
2. T & F = F
3. T | T = T
4. T | F = T

## Digression: `>`, `<`, `<=`, `>=`, `==`, `!=`

`x > y` If x is larger than y return `TRUE`.

`x < y` If x is smaller than y return `TRUE`.

`x <= y` If x is smaller or equal to y return `TRUE`.

`x >= y` If x is larger or equal to y return `TRUE`.

`x == y` If x is equal to y return `TRUE`.

`x != y` If x is unequal to y return `TRUE`.

## Recoding variables

- Create a dummy variable indicating a domestic car

```{r recode, eval=TRUE}
df$domestic <- FALSE
df$domestic[which(df$foreign == 'Domestic')] <- TRUE 
```

- Alternatively

```{r recode2, eval=TRUE}
df$domestic <- df$foreign == 'Domestic'
```

- The value TRUE is copied into all observations of the variable for which `foreign == 'Domestic'`
- Note the double equation sign `==`
- Note the quotation signs around 'Domestic'

## Recoding variables

- Create a variable which contains the price of domestic cars

```{r recode3, eval=TRUE}
df$price_domestic <- as.numeric(NA)

df$price_domestic[which(df$domestic == TRUE)] <- df$price

df$price_domestic[which(df$domestic == TRUE)] <-
  df$price[which(df$domestic == TRUE)]
```

- The vector on the RHS needs to be shorter or equal to the length of the vector on the LHS

## Why `stringsAsFactor = F`?

```{r stringsasfactors1}
df <- read.csv('data/BundestagForecastReplicationData.csv')
df$chancellor[15:16] <- 'Gerhard Schröder'
```

```{r stringsasfactors1_hidden, echo=FALSE, eval=TRUE}
df <- read.csv('../data/BundestagForecastReplicationData.csv')
df$chancellor[15:16] <- 'Gerhard Schröder'
```


```{r stringsasfactors2}
df <- read.csv('data/BundestagForecastReplicationData.csv',
               stringsAsFactors = F)
df$chancellor[15:16] <- 'Gerhard Schröder'
```
```{r stringsasfactors2_hidden, echo=FALSE, eval=TRUE}
df <- read.csv('../data/BundestagForecastReplicationData.csv',
               stringsAsFactors = F)
df$chancellor[15:16] <- 'Gerhard Schröder'
```

## Saving and loading objects

- R allows you to save any kind of object, not just datasets to the hard drive.

```{r save}
a <- 1
b <- c(T, F)

save(a, b, file = 'data/data.RData')
```
```{r save_hidden, eval=TRUE, echo=TRUE}
a <- 1
b <- c(T, F)

save(a, b, file = '../data/data.RData')
```

## Saving and loading objects

```{r load}
load('data/data.RData')

a
b
```
```{r load_hidden, eval=TRUE, echo=FALSE}
load('../data/data.RData')

a
b
```

- Load puts the objects `a` and `b` into R's memory

# Hands-on I

## Hands-on I

`hands-on/02_datamanipulation/hands-on1.R`